Liveness-Based Garbage Collection for Lazy Languages

### 2016-04-19 paper-for-archive/paper.tex
Added Acknowledgement and arxiv version

\documentclass[9pt,nonatbib]{sigplanconf}
\usepackage{etex}
\usepackage{float}
\usepackage{mathrsfs}
%%\usepackage{lineno}
\usepackage{booktabs}
\usepackage{stmaryrd}
\usepackage{boxedminipage}
\usepackage[T1]{fontenc}
\usepackage{epsfig}
\usepackage{multirow}
\usepackage{url}
\usepackage[normalem]{ulem}
\usepackage{schemepgm}
\usepackage{graphicx}
\usepackage{graphpap}
\usepackage{tabularx}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{pstricks}
\usepackage{pst-text}
\usepackage{pst-node}
\usepackage{pst-tree}
\usepackage{pst-rel-points}
\usepackage{bcprules}
\usepackage{subfigure}
\usepackage[boxed]{algorithm2e}
\usepackage{makecell}
\usepackage{balance}
\usepackage[numbers,sort&compress]{natbib}
\usepackage{wrapfig}
\usepackage{spreadtab}

% to save space in lists
\usepackage{enumitem}
\setlist{nolistsep,leftmargin=*}

@@

\definecolor{Myblue}{rgb}{.2,0,1}
\definecolor{Myred}{rgb}{1,.2,.2}

@@

\newcommand{\oxlb}{\ensuremath{[\!|}}
\newcommand{\oxrb}{\ensuremath{|\!]}}
\newcommand{\cred}[1]{\psframebox[linestyle=none, fillcolor=lightgray,fillstyle=solid,framesep=0.5pt]{#1}}
%\newcommand{\deltacall}[3]{\delta_{#1}({#2},{#3})}
\newcommand{\scmin} {\mbox{\sf\em in}}
\newcommand{\scmnull}{\mbox{\sf\em null?}}
\newcommand{\scmpair}{\mbox{\sf\em pair?}}
\newcommand{\scmprim}{\ensuremath{\mathsf{+}}}
\newcommand{\sh}[1]{{\colorbox{gray!20}{\framebox{$#1$}}}}
\def\myvec{\mathaccent"017E } 
\newcommand{\stk}{{S}}      
\newcommand{\ID}{\mbox{$\mathbf{ id}$}}
\newcommand{\bang}{\mbox{\sc bang}}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{lemma}[theorem]{Lemma}
\begin{document}

@@

% You need the command \numberofauthors to handle the 'placement
% and alignment' of the authors beneath the title.
%
% For aesthetic reasons, we recommend 'three authors at a time'
% i.e. three 'name/affiliation blocks' be placed beneath the title.
%
% NOTE: You are NOT restricted in how many 'rows' of
% "name/affiliations" may appear. We just ask that you restrict
% the number of 'columns' to three.
%
% Because of the available 'opening page real-estate'
% we ask you to refrain from putting more than six authors
% (two rows with three columns) beneath the article title.
% More than six makes the first-page appear very cluttered indeed.
%
% Use the \alignauthor commands to handle the names
% and affiliations for an 'aesthetic maximum' of six authors.
% Add names, affiliations, addresses for
% the seventh etc. author(s) as the argument for the
% \additionalauthors command.
% These 'additional authors' will be output/set for you
% without further effort on your part as the last section in
% the body of your article BEFORE References or any Appendices.
%\cmt{{
%
%\numberofauthors{3} %  in this sample file, there are a *total*
% of EIGHT authors. SIX appear on the 'first-page' (for formatting
% reasons) and the remaining two appear in the \additionalauthors
%section.
%
%\author{
% 1st. author
\authorinfo {Prasanna Kumar. K} 
       {IIT Bombay, Mumbai 400076, India}
       {prasannak@cse.iitb.ac.in}
% 2nd. author
\authorinfo {Amitabha Sanyal} 
       {IIT Bombay, Mumbai 400076, India}
       {as@cse.iitb.ac.in}
% 3rd. author
\authorinfo {Amey Karkare}
       {IIT Kanpur, Kanpur 208016, India}
       {karkare@cse.iitk.ac.in}
\maketitle

@@

Experiments    with   a    single    generation   copying    collector
(Section~\ref{sec:experiments})   confirm  the   expected  performance
benefits.  Liveness-based collection results in an increase in garbage
reclaimed.  As  a consequence, there is  a reduction in the  number of
collections from 1.6X  to 23X and a decrease  in the minimum
memory requirement from  1X to 1198X.  As  an added benefit,
there is also a reduction in the overall execution time in 5 out of 12
benchmark programs.

%------------------------------------------------------------%
\subsection{Motivating example}
\label{sec:motiv}
\input{fig_motiv}   
%%
Figure~\ref{fig:mot-example}   shows  an   example
program and the state of the heap at the program point $\pi$\footnote{We write
$\pi\!:\!e$ to  associate a  label $\pi$ with  the program  point just
before  the  expression  $e$. The label is not part of the syntax.},  i.e., just  before  the  evaluation  of
$(\length\ \pz)$.  The heap is represented  by a graph in which a node
either  represents  atomic  values   ($\NIL$,  integers  etc.),  or  a
\CONS\  cell  containing  $\CAR$  and  $\CDR$  fields,  or  a  closure
(represented  by  shaded  clouds).   Edges   in  the  graph  are  {\em
  references} and represent values of variables or fields.  The figure
shows the  lists \px\  and \pz\  partially evaluated  due to  the
\SIF\ condition 
(\NULLQ~(\CAR~\pz)).
The edges  shown by thick  arrows are those  which are live  at $\pi$.
%% Only those  cell which have a future use  should be preserved
%% during garbage collection, all other
%% cells can be  reclaimed.
Thus if a GC takes  place at $\pi$ with the heap shown
in Figure~\ref{fig:mot-example}(b),  a liveness-based  collector (LGC)
will preserve only  the cells referenced by $\pz$, and  the live cells
constituting the  closure referenced by $(\CDR~\pz)$.   In contrast, a
reachability-based collector (RGC) will preserve all cells.

In  this work  we show  that  static analysis  of heap  data can  help
garbage  collectors   in  reclaiming   more  garbage.    The  specific
contributions of this paper are:
\added{
\begin{itemize}
\item {\color{Myblue}We  propose an interprocedural liveness  based GC
  scheme  for a  lazy first  order functional  language and  prove its
  correctness.  To the  best of our knowledge, this is  the first work
  that uses  the results  of an  interprocedural liveness  analysis to
  garbage    collect    both     evaluated    data    and    closures.
  Thomas~\cite{Thomas19951} describes a  copying garbage collector for
  the Three  Instruction Machine (TIM)~\cite{Fairbairn1987}  that only
  preserves live closures  in a function's environment  (also called a
  frame). However, in the absence of  details, it is not clear whether
  a) the  scope of the  method is  interprocedural, and b)  it handles
  algebraic datatypes like lists (the original design of TIM did not).
  All  other  previous attempts~\cite{shaham01heap,  ran.shaham-sas03,
    shaham02estimating,  asati14lgc, karkare06effectiveness}  involved
  either imperative or eager functional languages.}
%% The novelty of our approach is in handling of
%%   liveness  of closures,  an  issue  that arises  due  to laziness  at
%%   runtime.}

@@

\item The  analysis results  in a set  of context-free  grammars along
  with  a fixed  set  of non-context-free  productions.  The  decision
  whether to copy a cell during  GC translates to a membership problem
  for such grammars.  Earlier, this decision problem was assumed to be
  undecidable, and the context-free grammar was over approximated by a
  regular  grammar.   In  this  paper,  we  provide  a  proof  of  the
  undecidability of this problem.

@@

\begin{figure}[t]\footnotesize
\renewcommand{\arraystretch}{0.9}
\begin{eqnarray*}
   p \in \mathit{Prog} & \!\!\!::=\!\!\! & d_1 \ldots d_n \,\,\,\, e_\mainpgm
   \hspace{5em} \,\,\,\,\,\,\,\,\,\; \mbox{\em --- program}\\
    \mathit{df} \in Fdef & \!\!\!::=\!\!\! & (\DEFINE\,\, (f\,\, x_1 \,\, \ldots
\,\,x_n)\,\,
    e)
    \hspace{0.2em} \ \ \ \ \ \ \ \ \  \mbox{\em --- function def} \\
e \in \mathit{Expr} & \!\!\!::=\!\!\! &
\left\{\begin{array}{@{}ll@{\hspace{1em}}l}
       (\SIF\,\, x\,\, e_1\,\, e_2) && \!\!\!\mbox{\em --- conditional} \\
       (\LET\,\, x \leftarrow s\,\, \IN\,\, e) &&\!\!\! \mbox{\em --- let
binding} \\
       (\SRETURN\,\, x) && \!\!\!\mbox{\em --- return from function}
    \end{array}\right. \\
s \in \mathit{App} & \!\!\!::=\!\!\!  &
\left\{\begin{array}{@{}l@{\hspace{1.2em}}l}
       k & \mbox{\em --- constant (numeric or $\NIL$)}\\
       (\CONS\,\, x_1\,\, x_2) & \mbox{\em --- constructor} \\
       (\CAR\,\, x) & \\
       (\CDR\,\, x) & \mbox{\em --- selectors} \\
       (\NULLQ\,\, x) & \\
       (\PRIM\,\, x_1\,\, x_2) & \mbox{\em ---  tester/generic arithmetic} \\
%%       (\ID\,\, x) & \mbox{\em ---  identity function (for inlining)} \\
       (f\,\, x_1\,\,\ldots\,\, x_n) & \mbox{\em --- function application}
    \end{array}\right.
\end{eqnarray*}
\vspace*{-6pt}
  \caption{The syntax of our language}\label{fig:lang-syntax}
%\figrule
\normalsize
\vspace*{-6pt}
\end{figure}

@@

$\heap(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ & $\rho,\, S,\, \heap,\, (+~x~y)
\rightsquigarrow
\rho',\, (\rho, \rho(x), (+~x~y))\!:\!S,\, \heap,\, s$      &
{\sc prim-1-clo} \\
\hline
$\heap(\rho(y)) \mbox{ is } \langle s, \rho'\rangle $ & $\rho,\, S,\, \heap,\, (+~x~y)
\rightsquigarrow
\rho', (\rho,\, \rho(y),\, (+~x~y))\!:\!S,\, \heap,\, s$      &
{\sc prim-2-clo} \\
\hline
{$\mathit{f}~\mbox{defined as}$
$~(\DEFINE~(f~\myvec{y})~e_{\mathit{f}})$}  & $\rho,\, S,\, \heap,\,
(f~\myvec{x})  \rightsquigarrow
[\myvec{y} \mapsto \rho(\myvec{x})],\, S,\, \heap,\, e_{\mathit{f}}$      &
{\sc funcall} \\
\hline
$\ell$ is a new location& {$\rho,\, S,\, \heap,\, (\LET~x\leftarrow s~\IN~e)
  \rightsquigarrow$
$\rho\oplus[x \mapsto \ell],\, S,\, \heap[\ell := \langle s,
    \lfloor\rho\rfloor_{FV(s)} %% \oplus [x \mapsto  \ell] -- NO LAZY LETS
    \rangle],\, e$} &
{\sc let} \\
\hline
$\heap(\rho(x)) \ne 0$ & $\rho,\, S,\, \heap,\, (\SIF~x~e_1~e_2)   \rightsquigarrow
\rho,\, S,\, \heap,\,  e_1$ & {\sc if-true} \\
\hline
$\heap(\rho(x)) = 0$ & $\rho,\, S,\, \heap,\, (\SIF~x~e_1~e_2)   \rightsquigarrow
\rho,\, S, \,\heap, \, e_2$ & {\sc if-false} \\
\hline
$\heap(\rho(x)) = \langle s, \rho' \rangle $ & {$\rho,\, S,\, \heap,\,
  (\SIF~x~e_1~e_2)   \rightsquigarrow
\rho',\, (\rho, \rho(x), (\SIF~x~e_1~e_2))\!:\!S,\, \heap, \, s$}
&
{\sc if-clo} \\
\hline
{$\heap(\rho(x))~\mbox{is in WHNF with value}~v$}& $\rho,\, (\rho', \ell,
e)\!:\!S,\, \heap,\,
(\SRETURN~x)  \rightsquigarrow \rho',\, S,\, \heap[\ell := v],\, e$ &
{\sc return-whnf}\\
\hline
$\heap(\rho(x)) = \langle s, \rho' \rangle $ & {$\rho,\, S,\, \heap,\, (\SRETURN~x)
  \rightsquigarrow$
$\rho',\, (\rho, \rho(x), (\SRETURN~x))\!:\!S,\, \heap,\,  s$} &
{\sc return-clo} \\
\hline
\end{tabular}
\caption{A small-step semantics for the language. \label{fig:lang-semantics}}
\end{center}
%}\comment
\vspace*{-16pt}
\end{figure*}

@@

\subsection{Liveness analysis for lazy languages}
\label{sec:liveness-analysis}
\input{example-to-illustrate-liveness}

@@

\input{fig-liveness}
Figure~\ref{fig:live-judge}  describes  our  analysis  which  has  two
parts. The  function $\mathit{ref}$,  takes an  application $s$  and a
demand $\sigma$ and returns the incremental liveness generated for the
free variables of $s$ due to  the application.  This will be consulted
during GC  while exploring the heap  starting from the
closure  variables.  The function $\mathcal{L}$  uses  $\mathit{ref}$ to  propagate
liveness across expressions.

@@

\begin{figure*}[t!]
\begin{center}\footnotesize
\renewcommand{\arraystretch}{1.2}
\begin{tabular}{|@{}c@{\ }|@{\ }c@{}|@{\ }c@{\ }|}
\hline
Premise & Transition & Rule name \\ 
\hline
\hline 
\makecell[t]{ $\rho(x) \mbox{ is } \bot$} & $\rho, S,
  \heap, (\CAR~x), \cred{\sigma} \rightsquigarrow \bang$ &
{\sc car-bang} 
\\
\hline
\makecell[t]{$\heap(\rho(x)) \mbox{ is } (\langle s, \rho'\rangle, d)$} & \makecell[t]{$\rho,\, S,\,  \heap,\,
(\CAR~x), \cred{\sigma} \rightsquigarrow$ $ \rho', \,(\rho, addr(\langle
s, \rho'\rangle), (\CAR~x),\cred{\sigma} )\!:\!S, \heap,\, s, \, \cred{\sigma}$ }     &
{\sc car-1-clo} 
\\
\hline
\makecell[t]{ $\heap(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$} & 
\makecell[t]{ $\rho, S, \heap, (\CAR~x), \cred{\sigma}
  \rightsquigarrow$ 
  $\rho', (\rho, \rho(x), (\CAR~x), \cred{\sigma})\!:\!S, \heap, s,\cred{\renewcommand{\arraystretch}{1}\left\{
    \begin{array}{lr}
      {\emptyset}&\mbox{if } \sigma = \emptyset \\       
      {\{\epsilon\} \cup \acar\sigma }&\mbox{otherwise}
    \end{array}\right.
  }$   }      
&
{\sc car-clo}\\
\hline
\makecell[t]{$\cred{GC(\rho_1, S_1, \heap_1, (\LET~x\leftarrow
    s~\IN~e), \sigma) = (\rho, S, \heap)}$,\\$\ell$ is a new location}& \makecell[t]{$\rho, S, \heap, (\LET~x\leftarrow
  s~\IN~e), \cred{\sigma}$  $ \rightsquigarrow \rho\oplus[x
    \mapsto \ell], S, \heap[\ell := \langle s, \lfloor\rho\rfloor_{FV(s)}, \sigma_x\rangle], e, \cred{\sigma}$ \\
    where $\sigma_x\ =\  \lfloor\mathcal{L}(e,\sigma,\Lfonly)\rfloor_{\{x\}}$} &
{\sc let} \\ 
\hline
\end{tabular}
\end{center}
\caption{Minefield  semantics for  \CAR\   and \LET. Complete minefield semantics is provided in~\cite{TR16-1}.
  \warning{Add a note mentioning that we handle the issue of requiring
    a      specific     demand      by      introducing     a      new
    symbol}\label{fig:minefield-semantics-for-some}}
\vspace*{-2mm}
\end{figure*}

@@

\noindent To set up the minefield semantics, we follow these steps:
\begin{enumerate}

@@

%%     For each
%%     location $\ell$,  we set $\heap(\ell)$  to $\bot$ iff for each
%%     environment $\rho$ in either  the  current state or in a
%%     continuation frame, for each $x \in domain(\rho)$ and each forward
%%     access
%%     path $\alpha$, it is not the case that $x.\alpha \in \Lv$ and $\heap\llbracket\rho(x), \alpha
%%     \rrbracket = \ell$. Here 
%% %% $x.\alpha$  is  live  and
%% %%     corresponds  to   $\ell$ in \heap.  More  formally,  there   are  no  $x\in
%% %%     domain(\rho)$ and  $\alpha$, such that  $\heap\llbracket\rho(x), \alpha
%% %%     \rrbracket = \ell$  and $x.\alpha \in \Lv$, where 
%% \Lv\ is the liveness environment  corresponding to $\rho$ mentioned in
%% step~\ref{gc-1}.
  \end{enumerate}
\end{enumerate}
Figure ~\ref{fig:minefield-semantics-for-some} shows some of the
minefield rules. As mentioned earlier, the transition for a \LET\ 
%%  $([\;]_\rho,([\;]_\rho,
%% \ans, (\print~\ans)):[\;]_{S}  , [\;]_{\heap}, (\mainpgm)$,
is preceded  by $GC(\ldots)$.  The details of the transition for the {\sc
  car-clo}  rule   is  also shown.  If    an
earlier  call to  $GC(\ldots)$  results  in $\rho(x)$  being
bound  to  $\bot$,  then  the  $\rightsquigarrow$  step
enters the  \bang\ state ({\sc  car-bang}). Otherwise
the transition is similar to the earlier {\sc car-clo}
rule.

>>>>

\documentclass[9pt,preprint,nonatbib]{sigplanconf}
\usepackage{balance}
\usepackage{float}
\usepackage{mathrsfs}
%%\usepackage{lineno}
\usepackage{booktabs}
\usepackage{stmaryrd}
\usepackage{etex}
\usepackage{boxedminipage}
\usepackage[T1]{fontenc}
\usepackage{epsfig}
\usepackage{multirow}
\usepackage{url}
\usepackage[normalem]{ulem}
\usepackage{schemepgm}
\usepackage{graphicx}
\usepackage{graphpap}
\usepackage{tabularx}
\usepackage{amssymb}
\usepackage{amsmath}
%\usepackage[pdf]{pstricks}
\usepackage{pst-text}
\usepackage{pst-node}
\usepackage{pst-tree}
%\usepackage{pst-rel-points}
\usepackage{bcprules}
\usepackage{subfigure}
\usepackage[boxed]{algorithm2e}
\usepackage{makecell}

% to save space in lists
\usepackage{enumitem}
\setlist{nolistsep}

@@

%\definecolor{Myblue}{rgb}{.2,0,1}
%\definecolor{Myred}{rgb}{1,.2,.2}

@@

\newcommand{\oxlb}{\ensuremath{[\!|}}
\newcommand{\oxrb}{\ensuremath{|\!]}}
%\newcommand{\cred}[1]{\psframebox[linestyle=none, fillcolor=lightgray,fillstyle=solid,framesep=0.5pt]{\ensuremath{#1}}}
\newcommand{\cred}[1]{\colorbox{lightgray}{\ensuremath{#1}}}
%\newcommand{\deltacall}[3]{\delta_{#1}({#2},{#3})}
\newcommand{\scmin} {\mbox{\sf\em in}}
\newcommand{\scmnull}{\mbox{\sf\em null?}}
\newcommand{\scmpair}{\mbox{\sf\em pair?}}
\newcommand{\scmprim}{\ensuremath{\mathsf{+}}}
\newcommand{\sh}[1]{{\colorbox{gray!20}{\framebox{$#1$}}}}
\def\myvec{\mathaccent"017E } 
\newcommand{\stk}{{S}}      
\newcommand{\ID}{\mbox{$\mathbf{ id}$}}
\newcommand{\bang}{\mbox{\sc bang}}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{lemma}[theorem]{Lemma}
\begin{document}

@@

% You need the command \numberofauthors to handle the 'placement
% and alignment' of the authors beneath the title.
%
% For aesthetic reasons, we recommend 'three authors at a time'
% i.e. three 'name/affiliation blocks' be placed beneath the title.
%
% NOTE: You are NOT restricted in how many 'rows' of
% "name/affiliations" may appear. We just ask that you restrict
% the number of 'columns' to three.
%
% Because of the available 'opening page real-estate'
% we ask you to refrain from putting more than six authors
% (two rows with three columns) beneath the article title.
% More than six makes the first-page appear very cluttered indeed.
%
% Use the \alignauthor commands to handle the names
% and affiliations for an 'aesthetic maximum' of six authors.
% Add names, affiliations, addresses for
% the seventh etc. author(s) as the argument for the
% \additionalauthors command.
% These 'additional authors' will be output/set for you
% without further effort on your part as the last section in
% the body of your article BEFORE References or any Appendices.
%\cmt{{
%
%\numberofauthors{3} %  in this sample file, there are a *total*
% of EIGHT authors. SIX appear on the 'first-page' (for formatting
% reasons) and the remaining two appear in the \additionalauthors
%section.
%
%\author{
% 1st. author
\authorinfo {Prasanna Kumar. K} 
       {IIT Bombay, Mumbai 400076, India}
       {prasannak@cse.iitb.ac.in}
% 2nd. author
\authorinfo {Amitabha Sanyal} 
       {IIT Bombay, Mumbai 400076, India}
       {as@cse.iitb.ac.in}
% 3rd. author
\authorinfo {Amey Karkare}
       {IIT Kanpur, Kanpur 208016, India}
       {karkare@cse.iitk.ac.in}

\maketitle

@@

Experiments    with   a    single    generation   copying    collector
(Section~\ref{sec:experiments})   confirm  the   expected  performance
benefits.  Liveness-based collection results in an increase in garbage
reclaimed.  As  a consequence, there is  a reduction in the  number of
collections from 1.6X  to 23X and a decrease  in the minimum
memory requirement from  1X to 1389X.  As  an added benefit,
there is also a reduction in the overall execution time in 5 out of 12
benchmark programs.

%------------------------------------------------------------%
\subsection{Motivating example}
\label{sec:motiv}

\begin{figure}[t!]
  \includegraphics[width=\columnwidth]{fig_motiv.pdf}
  \caption{Example Program and its Memory Graph}\label{fig:mot-example}
\end{figure}

  Figure~\ref{fig:mot-example}   shows  an   example
program and the state of the heap at the program point $\pi$\footnote{We write
$\pi\!:\!e$ to  associate a  label $\pi$ with  the program  point just
before  the  expression  $e$. The label is not part of the syntax.},  i.e., just  before  the  evaluation  of
$(\length\ \pz)$.  The heap is represented  by a graph in which a node
either  represents  atomic  values   ($\NIL$,  integers  etc.),  or  a
\CONS\  cell  containing  $\CAR$  and  $\CDR$  fields,  or  a  closure
(represented  by  shaded  clouds).   Edges   in  the  graph  are  {\em
  references} and represent values of variables or fields.  The figure
shows the  lists \px\  and \pz\  partially evaluated  due to  the
\SIF\ condition 
(\NULLQ~(\CAR~\pz)).
The edges  shown by thick  arrows are those  which are live  at $\pi$.
%% Only those  cell which have a future use  should be preserved
%% during garbage collection, all other
%% cells can be  reclaimed.
Thus if a GC takes  place at $\pi$ with the heap shown
in Figure~\ref{fig:mot-example}(b),  a liveness-based  collector (LGC)
will preserve only  the cells referenced by $\pz$, and  the live cells
constituting the  closure referenced by $(\CDR~\pz)$.   In contrast, a
reachability-based collector (RGC) will preserve all cells.

In  this work  we show  that  static analysis  of heap  data can  help
garbage  collectors   in  reclaiming   more  garbage.    The  specific
contributions of this paper are:
\added{
\begin{itemize}
\item We  propose a liveness  based GC scheme  for a lazy  first order
  functional language and  prove its correctness.  To the  best of our
  knowledge,  this is  the first  work on  liveness-based GC  for lazy
  languages.       All      previous      attempts~\cite{shaham01heap,
    ran.shaham-sas03,          shaham02estimating,         asati14lgc,
    karkare06effectiveness} have  involved either imperative  or eager
  functional languages.  The novelty of our approach is in handling of
  liveness  of closures,  an  issue  that arises  due  to laziness  at
  runtime.

@@

\item The  analysis results  in a set  of context-free  grammars along
  with  a fixed  set  of non-context-free  productions.  The  decision
  whether to copy a cell during  GC translates to a membership problem
  for such  grammars.  Earlier research assumed  the undecidability of
  this membership question and  used an over-approximation to overcome
  it.  In this paper, we provide  a formal proof of the undecidability
  of this problem.

@@

\begin{figure}[t]\footnotesize
\renewcommand{\arraystretch}{0.9}
\begin{eqnarray*}
   p \in \mathit{Prog} & \!\!\!::=\!\!\! & d_1 \ldots d_n \,\,\,\, e_\mainpgm
   \hspace{5em} \,\,\,\,\,\,\,\,\,\; \mbox{\em --- program}\\
    \mathit{df} \in Fdef & \!\!\!::=\!\!\! & (\DEFINE\,\, (f\,\, x_1 \,\, \ldots
\,\,x_n)\,\,
    e)
    \hspace{0.2em} \ \ \ \ \ \ \ \ \  \mbox{\em --- function def} \\
e \in \mathit{Expr} & \!\!\!::=\!\!\! &
\left\{\begin{array}{@{}ll@{\hspace{1em}}l}
       (\SIF\,\, x\,\, e_1\,\, e_2) && \!\!\!\mbox{\em --- conditional} \\
       (\LET\,\, x \leftarrow s\,\, \IN\,\, e) &&\!\!\! \mbox{\em --- let
binding} \\
       (\SRETURN\,\, x) && \!\!\!\mbox{\em --- return from function}
    \end{array}\right. \\
s \in \mathit{App} & \!\!\!::=\!\!\!  &
\left\{\begin{array}{@{}l@{\hspace{1.2em}}l}
       k & \mbox{\em --- constant (numeric or $\NIL$)}\\
       (\CONS\,\, x_1\,\, x_2) & \mbox{\em --- constructor} \\
       (\CAR\,\, x) & \\
       (\CDR\,\, x) & \mbox{\em --- selectors} \\
       (\NULLQ\,\, x) & \\
       (\PRIM\,\, x_1\,\, x_2) & \mbox{\em ---  tester/generic arithmetic} \\
%%       (\ID\,\, x) & \mbox{\em ---  identity function (for inlining)} \\
       (f\,\, x_1\,\,\ldots\,\, x_n) & \mbox{\em --- function application}
    \end{array}\right.
\end{eqnarray*}
%\vspace*{-6pt}
  \caption{The syntax of our language}\label{fig:lang-syntax}
%\figrule
\normalsize
%\vspace*{-6pt}
\end{figure}

@@

$\heap(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ & $\rho,\, S,\, \heap,\, (+~x~y)
\rightsquigarrow
\rho',\, (\rho, \rho(x), (+~x~y))\!:\!S,\, \heap,\, s$      &
{\sc prim-1-clo} \\
\hline
$\heap(\rho(y)) \mbox{ is } \langle s, \rho'\rangle $ & $\rho,\, S,\, \heap,\, (+~x~y)
\rightsquigarrow
\rho', (\rho,\, \rho(y),\, (+~x~y))\!:\!S,\, \heap,\, s$      &
{\sc prim-2-clo} \\
\hline
{$\mathit{f}~\mbox{defined as}$
$~(\DEFINE~(f~\myvec{y})~e_{\mathit{f}})$}  & $\rho,\, S,\, \heap,\,
(f~\myvec{x})  \rightsquigarrow
[\myvec{y} \mapsto \rho(\myvec{x})],\, S,\, \heap,\, e_{\mathit{f}}$      &
{\sc funcall} \\
\hline
$\ell$ is a new location& {$\rho,\, S,\, \heap,\, (\LET~x\leftarrow s~\IN~e)
  \rightsquigarrow$
$\rho\oplus[x \mapsto \ell],\, S,\, \heap[\ell := \langle s,
    \lfloor\rho\rfloor_{FV(s)} %% \oplus [x \mapsto  \ell] -- NO LAZY LETS
    \rangle],\, e$} &
{\sc let} \\
\hline
$\heap(\rho(x)) \ne 0$ & $\rho,\, S,\, \heap,\, (\SIF~x~e_1~e_2)   \rightsquigarrow
\rho,\, S,\, \heap,\,  e_1$ & {\sc if-true} \\
\hline
$\heap(\rho(x)) = 0$ & $\rho,\, S,\, \heap,\, (\SIF~x~e_1~e_2)   \rightsquigarrow
\rho,\, S, \,\heap, \, e_2$ & {\sc if-false} \\
\hline
$\heap(\rho(x)) = \langle s, \rho' \rangle $ & {$\rho,\, S,\, \heap,\,
  (\SIF~x~e_1~e_2)   \rightsquigarrow
\rho',\, (\rho, \rho(x), (\SIF~x~e_1~e_2))\!:\!S,\, \heap, \, s$}
&
{\sc if-clo} \\
\hline
{$\heap(\rho(x))~\mbox{is in WHNF with value}~v$}& $\rho,\, (\rho', \ell,
e)\!:\!S,\, \heap,\,
(\SRETURN~x)  \rightsquigarrow \rho',\, S,\, \heap[\ell := v],\, e$ &
{\sc return-whnf}\\
\hline
$\heap(\rho(x)) = \langle s, \rho' \rangle $ & {$\rho,\, S,\, \heap,\, (\SRETURN~x)
  \rightsquigarrow$
$\rho',\, (\rho, \rho(x), (\SRETURN~x))\!:\!S,\, \heap,\,  s$} &
{\sc return-clo} \\
\hline
\end{tabular}
\caption{A small-step semantics for the language. \label{fig:lang-semantics}}
\end{center}
%}\comment
%\vspace*{-16pt}
\end{figure*}

@@

\begin{figure}[t]
  \input{example-to-illustrate-liveness}  
  \caption{Example illustrating liveness of closures}\label{fig:lv-closure}
\end{figure}
\input{fig-liveness}

\subsection{Liveness analysis for lazy languages}
\label{sec:liveness-analysis}

@@

Figure~\ref{fig:live-judge}  describes  our  analysis  which  has  two
parts. The  function $\mathit{ref}$,  takes an  application $s$  and a
demand $\sigma$ and returns the incremental liveness generated for the
free variables of $s$ due to  the application.  This will be consulted
during GC  while exploring the heap  starting from the
closure  variables.  The function $\mathcal{L}$  uses  $\mathit{ref}$ to  propagate
liveness across expressions.

@@

@@

\begin{figure*}[t!]
\begin{center}\footnotesize
\renewcommand{\arraystretch}{1.2}
\begin{tabular}{|@{}c@{}|@{}c@{}|@{}c@{}|}
\hline
Premise & Transition & Rule name \\ 
\hline
\hline 
\makecell[t]{ $\rho(x) \mbox{ is } \bot$} & $\rho, S,
  \heap, (\CAR~x), \cred{\sigma} \rightsquigarrow \bang$ &
{\sc car-bang} 
\\
\hline
\makecell[t]{$\heap(\rho(x)) \mbox{ is } (\langle s, \rho'\rangle, d)$} & \makecell[t]{$\rho,\, S,\,  \heap,\,
(\CAR~x), \cred{\sigma} \rightsquigarrow$ $ \rho', \,(\rho, addr(\langle
s, \rho'\rangle), (\CAR~x),\cred{\sigma} )\!:\!S, \heap,\, s, \, \cred{\sigma}$ }     &
{\sc car-1-clo} 
\\
\hline
\makecell[t]{ $\heap(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$} & 
\makecell[t]{ $\rho, S, \heap, (\CAR~x), \cred{\sigma}
  \rightsquigarrow$ 
  $\rho', (\rho, \rho(x), (\CAR~x), \cred{\sigma})\!:\!S, \heap, s,\cred{\renewcommand{\arraystretch}{1}\left\{
    \begin{array}{@{}lr@{}}
      {\emptyset}&\mbox{if } \sigma = \emptyset \\       
      {\{\epsilon\} \cup \acar\sigma }&\mbox{otherwise}
    \end{array}\right.
  }$   }      
&
{\sc car-clo}\\
\hline
\makecell[t]{$\cred{GC(\rho_1, S_1, \heap_1, (\LET~x\leftarrow
    s~\IN~e), \sigma) = (\rho, S, \heap)}$,\\$\ell$ is a new location}& \makecell[t]{$\rho, S, \heap, (\LET~x\leftarrow
  s~\IN~e), \cred{\sigma}$  $ \rightsquigarrow \rho\oplus[x
    \mapsto \ell], S, \heap[\ell := \langle s, \lfloor\rho\rfloor_{FV(s)}, \sigma_x\rangle], e, \cred{\sigma}$ \\
    where $\sigma_x\ =\  \lfloor\mathcal{L}(e,\sigma,\Lfonly)\rfloor_{\{x\}}$} &
{\sc let} \\ 
\hline
\end{tabular}
\end{center}
\caption{Minefield  semantics for  \CAR\   and \LET.
  \warning{Add a note mentioning that we handle the issue of requiring
    a      specific     demand      by      introducing     a      new
    symbol}\label{fig:minefield-semantics-for-some}}
%\vspace*{-2mm}
\end{figure*}
\noindent To set up the minefield semantics, we follow these steps:
\begin{enumerate}

@@

%%     For each
%%     location $\ell$,  we set $\heap(\ell)$  to $\bot$ iff for each
%%     environment $\rho$ in either  the  current state or in a
%%     continuation frame, for each $x \in domain(\rho)$ and each forward
%%     access
%%     path $\alpha$, it is not the case that $x.\alpha \in \Lv$ and $\heap\llbracket\rho(x), \alpha
%%     \rrbracket = \ell$. Here 
%% %% $x.\alpha$  is  live  and
%% %%     corresponds  to   $\ell$ in \heap.  More  formally,  there   are  no  $x\in
%% %%     domain(\rho)$ and  $\alpha$, such that  $\heap\llbracket\rho(x), \alpha
%% %%     \rrbracket = \ell$  and $x.\alpha \in \Lv$, where 
%% \Lv\ is the liveness environment  corresponding to $\rho$ mentioned in
%% step~\ref{gc-1}.
  \end{enumerate}
\end{enumerate}
Figure ~\ref{fig:minefield-semantics-for-some} shows some of the
minefield rules. As mentioned earlier, the transition for a \LET\ 
%%  $([\;]_\rho,([\;]_\rho,
%% \ans, (\print~\ans)):[\;]_{S}  , [\;]_{\heap}, (\mainpgm)$,
is preceded  by $GC(\ldots)$.  The details of the transition for the {\sc
  car-clo}  rule   is  also shown.  If    an
earlier  call to  $GC(\ldots)$  results  in $\rho(x)$  being
bound  to  $\bot$,  then  the  $\rightsquigarrow$  step
enters the  \bang\ state ({\sc  car-bang}). Otherwise
the transition is similar to the earlier {\sc car-clo}
rule. The remaining rules for minefield semantics are given in Appendix~\ref{sec:minefield}.
